---
title: "Hands On 1"
author: "Regina Rodriguez Durant (regina.rodriguezdurant01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This practical focuses on data wrangling, exploratory data analysis, visualization, and critical comparison with published results using real biomedical data.

We will work with the publication [**Proteomic and Metabolomic Characterization of COVID-19 Patient Sera**](https://www.sciencedirect.com/science/article/pii/S0092867420306279)

# Exercise 1  

Download and load Supplementary Table 1 (“Additional Demographical and Baseline Characteristics of COVID-19 Patients and Control Groups”) from the supplementary materials of the publication.
 
## Exercise 1.1  

Reproduce Table 1 from the manuscript using the supplementary data. Match the reported summary statistics   as closely as possible.

Briefly discuss any discrepancies between your results and the published table, and provide possible explanations.

```{r}



```


## Exercise 1.2
Perform an exploratory data analysis (EDA) of the dataset. At a minimum, include:

- Descriptive statistics for key variables  
- Appropriate visualizations (e.g. distributions, group comparisons)  
- A brief written summary of notable patterns, anomalies, or data quality issues  


```{r}

```

# Exercise 2
## Exercise 2.1
Reproduce Supplementary Figure 1 using the data without modifying or removing any observations. Use the same variables and groupings as in the original figure. 
```{r}

```


## Exercise 2.2
Identify and handle outliers using an appropriate and well-justified method (e.g. removal, transformation, or robust statistics).

Reproduce Supplementary Figure 1 after outlier handling and discuss how and why the results change.
```{r}

```

# Exercise 3
Create a heatmap of the biomarker data. Include group and gender as annotation variables.

```{r}

```

# Conclusion

Write a concise conclusion (1 paragraphs) summarizing the main differences observed across the three patient groups based on the twelve clinical parameters. Support your statements with evidence from your analyses.

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
